---
title: "Introduction to EZMultitaper package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Intro_to_EZMultitaper}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(EZMultitaper)
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.align = "center",
    fig.width = 6,
    fig.height = 4
)
```

This is the introduction to how to use the EZMultitaper package.
It calculates the mean power matrix in a frequency range of intracranial electrocorticographic (iEEG) recordings of ictal events.

This mean power computation as well as the time statistics implemented in this package are based on the methods described in the paper 
by Li et al. (2021) <doi: 10.1038/s41593-021-00901-w>.
<a href="https://pubmed.ncbi.nlm.nih.gov/34354282/" style="color:blue; text-decoration:none;"> Mean Power in a frequency range</a>.

The package uses the multitaper spectrogram analysis described in the paper by Prerau et al. (2016) <doi:10.1152/physiol.00062.2015>
<a href="https://pubmed.ncbi.nlm.nih.gov/27927806/" style="color:blue; text-decoration:none;"> Multitaper Spectral Analysis</a>.


## Key definition of the mean power marker

Drug-resistant epilepsy (DRE) is a significant clinical challenge. Effective treatment requires the synthesis of large amounts of personalized clinical data with the goal of identifying the region within a patient’s brain from which seizures originate (the epileptogenic zone, or EZ). Current standard of care involves multidisciplinary group meetings and clinician consensus after reviewing patient data. This lack of a standardized methodology has led to variable surgical outcomes, measured by seizure freedom/reduction after surgery. Computational methods to identify the EZ have been developed to improve outcomes. 


## Load small ictal ieeg data for patient PT01

This example data corresponds to the first seizure of patient PT01 from the Multitaper Data Set. `pt01EcoG` is a matrix with 50001 columns (time points) and 14 rows (electrodes). The acquisition frequency is 1000Hz and the time around seizure onset is [-30:20]s. The rows names are the electrodes names. The column names is the time relative to seizure onset in (s). For more information about the data, you can use the `?pt01EcoG` command.

The patient underwent a surgical treatment and was seizure-free after the surgery. The electrodes located in the SOZ and the surgically removed area are marked in `sozIndex` attribute.
```{r ictal data}
data("pt01EcoG")
sozIndex <- attr(pt01EcoG, "sozIndex")

## create an epoch object
epoch <- Epoch(pt01EcoG)
#epoch
```

The function `visuIEEGData` can visualize the iEEG data. 
```{r}
visuIEEGData(epoch  = epoch)
```

## Compute Multitaper Matrix
The function `meanPowBaselineBand` computes the mean power matrix for the given iEEG time series in a given frequency band. The simplest form is `meanPowBaselineBand(epoch,windowParams, rangeBand)`.
```{r}
windowParams<-c(0.25,0.1)
betaBand<-c(13,30)
betaBandPow<-meanPowBaselineBand( epoch=epoch, fs=1000, windowParams=windowParams, rangeBand=betaBand)
```

## Plot functions for the mean power Matrix
The function `plotPowHeatmap` produce a heatmap of the mean power in a given frequency band and allows to visually check the correlation between the SOZ and sustained high mean power value. Time window around seizure onset [0:20]s, which includes the following electrodes: 

- Electrodes within the surgery region: `sozName`
- Electrodes outside of the surgery region: "MLT1", "MLT2", "MLT3", "MLT4" 

The parameter `sozIndex` accepts either electrode names or their indices in the mean power matrix. If `sozIndex` is provided, the function will display the `sozIndex` electrodes in blue in the top rows of the heatmap. If `sozIndex` is not provided, the function will simply generate a heatmap of the mean power matrix.

```{r}
plotPowBand<-plotPowHeatmap(pow=betaBandPow,sozIndex=sozIndex)
plotPowBand<-plotPowBand+ggplot2::ggtitle(("Mean beta power heatmap for patient pt01"))
plotPowBand
```


`plotPowDistribution` and `plotPowQuantile` are similar to the `plotPowHeatmap` function. They plot the mean power distribution and quantiles, respectively.
```{r, out.width="100%"}
## plot the mean power quantiles
plotbetaQuantile<-plotPowQuantile(pow = betaBandPow, sozIndex = sozIndex)
plotbetaQuantile<-plotbetaQuantile+ggplot2::ggtitle(("Pooled mean beta power quantiles for patient pt01"))
plotbetaQuantile
## plot the mean power distribution

```

```{r, out.width="100%"}
plotBetaDistr<-plotPowDistribution(pow =betaBandPow, sozIndex = sozIndex)
plotbetaDistr<-plotbetaDistr+ggplot2::ggtitle(("Pooled mean beta power distribution for patient pt01"))
plotBetaDistr
```


